### 2016-03-08 paper/icfp16-submission/gallery2.tex
fixed genWitness

\section{The Advantage Of Traces}
\label{sec:advantage-traces}
Having shown that \toolname produces small traces in an efficient
manner, let us discuss the explanatory power of traces compared with
typical type error messages. In this section we will present a
qualitative evaluation of \toolname's jump-compressed traces on a series
of examples drawn from real student programs in the \ucsdbench
dataset. We will show how \toolname's traces get to the heart of the
error by
%
\begin{enumerate}
\item highlighting the conflicting values that cause the program to get
  stuck, rather that blaming a single one,
\item showing the steps necessary to reach the stuck state, and
\item not assuming that a function is correct just because it type-checks.
\end{enumerate}
%
For each example we will present the code, the type error produced by
the \ocaml compiler, and a jump-compressed trace produced by \toolname.

@@

First, we will examine a simple recursive function @sqsum@ that should
square each element of the input list and then compute the sum of the
result. In the following code snippets we will underline the source span
that \ocaml blames for the type error.
%
\begin{ecode}
  let rec sqsum xs = match xs with
    | [] -> 0
    | h::t -> __(sqsum t)__ @ (h * h)
\end{ecode}
%
Unfortunately the student has used the list-append operator |@| instead
of \texttt{+} to compute the sum. \ocaml blames the recursive call
\texttt{sqsum t} on line 3 for being an \texttt{int} instead of a
\texttt{list}.
%
\begin{verbatim}
  This expression has type
    int
  but an expression was expected of type
    'a list
\end{verbatim}
%
In contrast, \toolname produces a trace showing how the evaluation of
@sqsum [1]@ eventually gets stuck when it tries to evaluate
\texttt{ 0 }|@|\texttt{ 1}.
%
\begin{center}
  \includegraphics[height=125px]{sqsum.png}
\end{center}
%
Note how the eye is drawn to the entire stuck term as opposed to just
the left sub-term. This emphasizes the \emph{conflict} between @int@ and
@list@ rather than assuming one or the other is correct.

Next we consider a function @sumList@ that just sums the elements of the
input list, but has a more unfortunate error.
\begin{ecode}
  let rec sumList xs = match xs with
    | []    -> []
    | y::ys -> y + __sumList ys__
\end{ecode}
%
In this case the student has returned @[]@ in the base case instead of
@0@. Unfortunately, \ocaml deduces (incorrectly) from the base case that
@sumList@ must return a @list@, and then proceeds to blame the
\emph{recursive call} on line 3 for producing a @list@ instead of an
@int@.
%
\begin{verbatim}
  This expression has type
    'a list
  but an expression was expected of type
    int
\end{verbatim}
%
\toolname's jump compressed trace shows how the evaluation of
@sumList [1; 2]@ gets stuck when it tries to evaluate @2 + []@.
%
\begin{center}
  \includegraphics[height=125px]{sumlist.png}
\end{center}
%
This is essentially the same error that \ocaml produces, but the trace
clarifies immediately (via the third step) that the @[]@ is the result
of the recursive call @sumList []@, drawing attention to the incorrect
base case.

>>>>

\section{The Advantage Of Traces}\label{sec:advantage-traces}

Next, we present a \emph{qualitative} evaluation that compares
the explanations provided by \toolname's dynamic witnesses with
the static reports produced by the \ocaml\ compiler and \sherrloc,
a state-of-the-art fault localization approach~\cite{ZhangMyers}.
%
In particular, we illustrate, using a series of examples drawn
from student programs in the \ucsdbench\ dataset how \toolname's
jump-compressed traces can get to the heart of the error by
%
highlighting the conflicting values that cause the program to get
stuck, rather that blaming a single one,
%
showing the steps necessary to reach the stuck state, and
%
not assuming that a function is correct just because it type-checks.
%
For each example we will present
(1)~the code,
(2)~the error message returned \ocaml,
(3)~the error locations returned by \ocaml\ (underlined) and \sherrloc\ (in bold),
(4)~the jump-compressed trace produced by \toolname.

@@

\paragraph{Example: Recursion with Bad Operator}
The recursive function @sqsum@ should square each
element of the input list and then compute the sum
of the result.
%
\begin{ecode}
  let rec sqsum xs = match xs with
    | [] -> 0
    | h::t -> ==(sqsum t)== @ (h * h)
\end{ecode}
%
Unfortunately the student has used the list-append
operator |@| instead of \texttt{+} to compute the sum.
%
Both \ocaml\ and \sherrloc\ blame the \emph{wrong location},
namely the recursive call @sqsum t@ with the message
%
\begin{verbatim}
  This expression has type
    int
  but an expression was expected of type
    'a list
\end{verbatim}
%
\toolname\ produces the following trace showing how the evaluation of
@sqsum [1]@ gets stuck:
%
\begin{center}
  \includegraphics[height=125px]{sqsum.png}
\end{center}
%
The figure highlights the entire stuck term
(not just the recursive call), emphasizing
the \emph{conflict} between @int@ and @list@
rather than assuming one or the other is correct.

\paragraph{Example: Recursion with Bad Base Case}
%
The function @sumList@ is supposed to add up
the elements of its input list.
%
\begin{ecode}
  let rec sumList xs = match xs with
    | []    -> ==[]==
    | y::ys -> y + __sumList ys__
\end{ecode}
%
Here, the student has incorrectly returned @[]@
in the base case instead of @0@.
%
Unfortunately, \ocaml\ deduces (incorrectly)
from the base case that @sumList@ must return
a @list@, and then proceeds to blame the
\emph{recursive call} on line 3 for
producing a @list@ instead of an @int@.
%
\begin{verbatim}
  This expression has type
    'a list
  but an expression was expected of type
    int
\end{verbatim}
%
\toolname's jump compressed trace shows how the evaluation of
@sumList [1; 2]@ gets stuck when it tries to evaluate @2 + []@.
%
\begin{center}
  \includegraphics[height=125px]{sumlist.png}
\end{center}
%
This is essentially the same error that \ocaml produces, but the trace
clarifies immediately (via the third step) that the @[]@ is the result
of the recursive call @sumList []@, drawing attention to the incorrect
base case.
